---
title: "Tutorial 1"
author: "Jakub Janusz"
date: "2026-04-29"
output: html_document
---

# 1 Handling social network data in R

Loading packages

```{r}
#install.packages("network")
#install.packages("igraph")
#install.packages("reshape2")
#install.packages("tidyverse")

library(network)
library(igraph)
library(reshape2)
library(tidyverse)
```

## 1.1 Getting network data into R: the igraph package

### 1.1.1 From adjacency matrix to igraph object

Reading data from internet

```{r}
url1 <- "https://github.com/rensec/sasr08/raw/main/g_adj_matrix_simple.csv"
g_matrix <- read.csv(file = url1, header = FALSE)
g_matrix
```

\^ adjacency matrix - square matrix in which both the rows and the columns represent nodes, and the values in the cells indicate the status of the edges between these nodes

```{r}
class(g_matrix)
```

Transformation into matrix object

```{r}
g_matrix <- as.matrix(g_matrix)
g_matrix
```

\^matrix object - the data still have the *shape* of an adjacency matrix, but in R, it is now stored as an object of class ‘matrix’

Adding row- and column names for claritry (by simply numbering them):

```{r}
rownames(g_matrix) <- 1:nrow(g_matrix)
colnames(g_matrix) <- 1:ncol(g_matrix)
g_matrix
```

**QUESTION:** How many nodes are included in this matrix, and how many edges are there between these nodes?

5 nodes and 4 edges between them

Reading an attribute of the nodes:

```{r}
g_nodes_age <- read.csv(file = "https://github.com/rensec/sasr08/raw/main/g_nodes_age.csv")
g_nodes_age
```

Creating an `igraph` network object from our matrix

```{r}
g <- graph_from_adjacency_matrix(g_matrix)
class(g)

g
```

Creating network map

```{r}
plot(g)

```

Adding node attributes to the object

```{r}
g <- set_vertex_attr(g, 
                     name = "age",
                     value = g_nodes_age$age)
```

### 1.1.2 From edge list to igraph object

Loading the file with edge list into a data frame:

```{r}
g_elist <- read.csv(file = "https://github.com/rensec/sasr08/raw/main/g_elist.csv")
g_elist
```

Creating an `igraph` network object from the edge list:

```{r}
g_from_elist <- graph_from_data_frame(g_elist) 
plot(g_from_elist)
```

\^ Isolate (mr 5) lost :(((

Now we want to include isolante in the network object

```{r}
g_from_elist <- graph_from_data_frame(g_elist, vertices = g_nodes_age) 
plot(g_from_elist)
```

! it is important to always include a complete node list as well (unless you are certain that the edge list includes all nodes)!

**Assignment:** from the code snippets above, construct a complete “pipeline” to go from the edge list to the igraph object. Start from `g_elist`. Plot the resulting network. Store the code somewhere for later use.

```{r}
g_elist <- read.csv(file = "https://github.com/rensec/sasr08/raw/main/g_elist.csv")

g_from_elist <- graph_from_data_frame(g_elist, vertices = g_nodes_age) 
plot(g_from_elist)
```

### 1.1.3 From adjacency list to igraph object

Adjacency list

```{r}
g_adj_list <- read.csv(file = "https://github.com/rensec/sasr08/raw/main/g_adj_list.csv")
g_adj_list
```

The most convenient way to create a network object from these data, is to first transform them into an *edge list*

Making data long

```{r}
g_elist_from_alist <- g_adj_list
g_elist_from_alist <-  select(g_elist_from_alist, id, friend1:friend2) # keep only the network variables (and id)
g_elist_from_alist <-  melt(g_elist_from_alist, id.vars = "id") # from the reshape2 package
g_elist_from_alist
```

Using pipe operator to streamline the code

```{r}
g_elist_from_alist <- g_adj_list %>% 
  select(id, friend1:friend2) %>% # keep only the network variables (and id)
  melt( id.vars = "id") # from the reshape2 package
```

Same thing using tidyverse

```{r}
g_adj_list %>% # (we don't save it into an object as above this time, since we just want to show that the result is the same)
  select(id, friend1:friend2) %>% #keep only the network variables (and id)
  pivot_longer(c(friend1, friend2))
```

To create an edge list we subsequently drop all missing values on “value” and do some housekeeping:

```{r}
g_elist_from_alist <- g_elist_from_alist%>% 
  filter(!is.na(value)) %>% # drop the missings
  rename(from ="id", to = "value", sourcevar= "variable") %>% #just nice for interpretation
  relocate(to, .after=from) #move around the columns
g_elist_from_alist
```

**QUESTION:** Besides being useful for data manipulation procedures, keeping the information in `sourcevar` could also be important for more substantive reasons. Can you think of such a reason?

For when we want to account for the strength of relationships? friend 1 might indicate best friend = stronger tie.

now importing into network object

```{r}
g_from_alist <- graph_from_data_frame(g_elist_from_alist) #This is a proper igraph graph object
plot(g_from_alist)
```

correction to include isolate (+creating adjacency list)

```{r}
nodelist <- select(g_adj_list,id,age)
g_from_alist <- graph_from_data_frame(g_elist_from_alist, vertices = nodelist) #This is a proper igraph graph object
plot(g_from_alist)
```

## 1.2 The reverse direction: exporting from igraph objects

### 1.2.1 From igraph to adjacency matrix

```{r}
g_adj_matrix <- as_adjacency_matrix(g, sparse = FALSE)
g_adj_matrix

```

### 1.2.2 From igraph to edge list

```{r}
g_edgelist <- igraph::as_data_frame(g_from_elist, what = "edges") 
g_edgelist
```

the number 5 again...

```{r}
igraph::as_data_frame(g_from_elist, what = "vertices") 

```

### 1.2.3 From igraph to adjacency list

```{r}
edgelist_2 <- igraph::as_data_frame(g_from_alist, what = "edges")
nodelist_2 <- igraph::as_data_frame(g_from_alist, what = "vertices")
# Note the use of the "namespace" "igraph::" here, to indicate that we need the igraph function here, not the function with the same name from the dplyr/tidyverse package

d <- edgelist_2 %>%
  rename(id = "from") %>% 
  pivot_wider( id_cols = id, names_from = sourcevar, values_from = to ) %>% 
  merge(nodelist_2, by.x = "id", by.y = "name", all.y = TRUE ) %>% 
  relocate(age, .after = id) %>% 
  type_convert() #as_data_frame returns characters; this transforms it back to numeric
```

```{r}
d
```

## 1.3 Getting network data into R: the network package

```{r}
g_np <- as.network(g_elist_from_alist, matrix.type = "edgelist", vertices = nodelist)
g_np

```

```{r}
plot(g_np)

```

## 1.4 Modifying networks in igraph

Removing a node :((

```{r}
g_mod <- delete_vertices(g,2)
plot(g_mod)
```

Eliminating 21 year olds

```{r}
g_mod <- delete_vertices(g,which(V(g)$age == 21))
plot(g_mod)
```

Eliminating missing age

```{r}
g_mod <- delete_vertices(g,which(is.na(V(g)$age)))
plot(g_mod)
```

Adding node attributes

```{r}
V(g)$gender <- c("male", "female", "female", "other", "male") 
g
```

Functions that modify the entire graph

```{r}
g_rev <- reverse_edges(g)
plot(g_rev)
```

combining edges of two graphs

```{r}
g_union <- igraph::union(g, g_rev)
plot(g_union)


g_undir <- as_undirected(g)
plot(g_undir)
```

***QUESTION:*** What’s the difference?

in one we see that there are edges in two directions and in the second one we just can see that there are edges
